Fast Sorting on Modern Computers

### 2005-07-31 docs/tech_report/conclusions.tex
the grand reaming is almost complete

We found that algorithm S has a lower instruction count than algorithm N, but
that this does not necessarily translate to better performance. Our results show
also that the level 1 cache needs to be taken into account, that the extra
instruction in multi-mergesort make it slower than double tiled mergesort, and
that multi-mergesort has regular branch predictive access patterns during the
k-way merge, which can be taken advantage of by a 2-level adaptive predictor.

@@

We were able to reduce the number of cache misses considerably using a technique
adapted from LaMarca's tiled mergesort, and managed to make cache radixsort the
fastest sort in our tests. This is due to its low instruction count, few cache
misses and lack of branch mispredictions. Although we attempted to align it
using another technique from mergesort, we found this decreased its performance.

@@

LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
misses, instruction count and cycle count. Like LaMarca, we found that a four
heap lowered our instruction count. Our level 2 misses also concurred with
LaMarca, with only one difference: cache heapsort, being an out-of-place sort,
fills the cache fully with a smaller number of keys than base heapsort, which is
in place. LaMarca's results do not show this. Nor do they show a difference in
cache misses between these two sorts when the array does fit in the cache.
LaMarca's base heapsort have the characteristics of an out-of-place heapsort.
However, our results show that his implicit heap improvements have led to a
significantly faster heapsort.

@@

Where LaMarca found that using Sedgewicks insertion sort optimisation decreased
instruction count, we found the opposite. We also found that despite
improvements due to cache performance and instruction count, our base quicksort
performed faster than our tuned version. Finally, we found that although
changing multi-mergesort's sequential search to a binary search decreased the
instruction count, it greatly increased the number of branch mispredictions
incurred, and slowed the sort down accordingly.

@@

The best stable sort is radixsort. The best comparison-based stable sort is
double tiled mergesort. The best in-place sort is quicksort. The best in-place
stable sort is, sadly, insertion sort. The best elementary sort is also
insertion sort; as such it should be only used for very small data sets.  The
best sort for very large data sets is radixsort. The best overall sort was
quicksort, though radixsort is very close, and the radixsort considered here was
not optimized. It is expected that an optimized radixsort would beat quicksort
for smaller sets.
% TODO redo above paragraph

>>>>

We found that algorithm S has a lower instruction count than algorithm N, but
that this does not necessarily translate to better performance. Our results show
also that the level 1 cache needs to be taken into account, that the extra
instruction in multi-mergesort make it slower than double-aligned tiled
mergesort, and that multi-mergesort has regular branch predictive access
patterns during the k-way merge, which can be taken advantage of by a 2-level
adaptive predictor.

@@

We were able to reduce the number of cache misses considerably using a technique
adapted from LaMarca's tiled mergesort, and managed to make memory-tuned radixsort the
fastest sort in our tests. This is due to its low instruction count, few cache
misses and lack of branch mispredictions. Although we attempted to align it
using another technique from mergesort, we found this decreased its performance.

@@

LaMarca's heapsort graphs were equivalent to ours in the areas of level 1
misses, instruction count and cycle count. Like LaMarca, we found that a four
heap lowered our instruction count. Our level 2 misses also concurred with
LaMarca, with only one difference: memory-tuned heapsort, being an out-of-place sort,
fills the cache fully with a smaller number of keys than base heapsort, which is
in place. LaMarca's results do not show this. Nor do they show a difference in
cache misses between these two sorts when the array does fit in the cache.
LaMarca's base heapsort have the characteristics of an out-of-place heapsort.
However, our results show that his implicit heap improvements have led to a
significantly faster heapsort.

@@

Where LaMarca found that using Sedgewicks insertion sort optimisation decreased
instruction count, we found the opposite. We also found that despite
improvements due to cache performance and instruction count, our base quicksort
performed faster than our memory-tuned version. Finally, we found that although
changing multi-mergesort's sequential search to a binary search decreased the
instruction count, it greatly increased the number of branch mispredictions
incurred, and slowed the sort down accordingly.

@@

The best stable sort is radixsort. The best comparison-based stable sort is
double-aligned tiled mergesort. The best in-place sort is quicksort. The best in-place
stable sort is, sadly, insertion sort. The best elementary sort is also
insertion sort; as such it should be only used for very small data sets.  The
best sort for very large data sets is radixsort. The best overall sort was
quicksort, though radixsort is very close, and the radixsort considered here was
not optimized. It is expected that an optimized radixsort would beat quicksort
for smaller sets.
% TODO redo above paragraph

### 2005-07-31 docs/tech_report/heap.tex
done filename renaming inside files
the grand reaming is almost complete

\section{Cache Heapsort}
\label{cache heapsort}
LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
misses than Floyd's. The reason for this is simple. Consider the number of keys 
touched when a new key is added to the heap. In the case of Floyd's method, the
key being heapified is not related to the last key that was heapified. Reuse
only occurs upon reaching a much higher level of the heap. For large heaps, these
keys may be swapped out of the cache by this time. As a result, Floyd's method
has poor temporal locality.

The Repeated-Adds algorithm, however, displays excellent temporal locality. A
key has a 50\% chance of having the same parent, a 75\% of the same grandparent
and so on. The number of instructions, however, increases significantly, and the
sort becomes out-of-place. For this reason, the Repeated-Adds algorithm is used
for cache heapsort.

@@

LaMarca reports base heapsort has very poor temporal locality, and that the
cache heapsort will have half the cache misses of base heapsort.

@@

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. Cache heaport also
has a similar increase, though LaMarca's results show that these occur at the
same position in the graph, in other words, when th set size is the same. Our
results, however, show that this occurs earlier in cache heapsort, due to the
fact that it is an out of place sort.

Finally, though this is likely to be related the previous differences, base
heapsort performs slightly better for small sets than the cache heapsorts do in
the results presented here. LaMarca's results indicate that the cache heapsort
has better results the entire way through the cache. This may relate to the
associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
unoptimized algorithm would not suffer as much.

@@

\sixplot{base_heapsort_0.eps}{cache4_heapsort_0.eps}{cache_heapsort_0.eps}{cache_heapsort_1.eps}{}{}{base and cache heapsorts}

The results of the software predictor show the relation of the branches
in the creation and destruction of the heap. The base heapsort results in
Figure \ref{base and cache heapsorts-1} show that the time taken to create the
heap is not significant relative to the time taken to sort from the heap.

The most interesting thing is the relation between the predictablility of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
promotion}' in Figure \ref{base and cache heapsorts-1} (the 2-heap), with the
columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure \ref{base and
cache heapsorts-2} (the 4-heap) and Figures \ref{base and cache heapsorts-3} and
\ref{base and cache heapsorts-4} (the 8-heap).

@@

More importantly, the accuracy of the each
successive comparison increases in the same way as selection sort (see Section
\ref{selection branches logn}), becoming more predictable as the fanout
increases. For example, in the 8-heap, shown in Figures \ref{base and cache heapsorts-3}
and \ref{base and cache heapsorts-4},
the first comparison is predicted correctly 50\% of the
time, while the next is predicted correctly 61\% of the time, then 71\%, 77\%, and
so on in a curve, which is shown in Figure \ref{cache heapsort curve}.

\begin{figure}
\includegraphics{plots/curve.eps}
\caption{Predictability of branches in cache heapsort, with a fanout of eight}
\label{cache heapsort curve}
\end{figure}

>>>>

\section{Memory-tuned Heapsort}
\label{memory-tuned heapsort}
LaMarca notes that the Repeated-Adds algorithm has a much lower number of cache
misses than Floyd's. The reason for this is simple. Consider the number of keys 
touched when a new key is added to the heap. In the case of Floyd's method, the
key being heapified is not related to the last key that was heapified. Reuse
only occurs upon reaching a much higher level of the heap. For large heaps, these
keys may be swapped out of the cache by this time. As a result, Floyd's method
has poor temporal locality.

The Repeated-Adds algorithm, however, displays excellent temporal locality. A
key has a 50\% chance of having the same parent, a 75\% of the same grandparent
and so on. The number of instructions, however, increases significantly, and the
sort becomes out-of-place. For this reason, the Repeated-Adds algorithm is used
for memory-tuned heapsort.

@@

LaMarca reports base heapsort has very poor temporal locality, and that the
memory-tuned heapsort will have half the cache misses of base heapsort.

@@

Slight differences exist though. The level 2 cache miss results here indicate
that while still sorting in the cache, the number of misses in the in-place
heapsort is less than that of the out-of-place sorts. LaMarca, however, has the
same amount of misses in both cases. Similarly, the results here indicate that
base heapsort experiences a sharp increase in cache misses. Cache heaport also
has a similar increase, though LaMarca's results show that these occur at the
same position in the graph, in other words, when th set size is the same. Our
results, however, show that this occurs earlier in memory-tuned heapsort, due to the
fact that it is an out of place sort.

Finally, though this is likely to be related the previous differences, base
heapsort performs slightly better for small sets than the memory-tuned heapsort's do in
the results presented here. LaMarca's results indicate that the memory-tuned heapsort
has better results the entire way through the cache. This may relate to the
associativity: LaMarca's tests were performed on a DEC AlphaStation 250 with a
direct mapped cache. The Pentium 4's cache is 8-way associative, and so the
unoptimized algorithm would not suffer as much.

@@

\sixplot{base_heapsort_0.eps}{memory_tuned4_heapsort_0.eps}{memory_tuned_heapsort_0.eps}{memory_tuned_heapsort_1.eps}{}{}{base and memory-tuned heapsorts}

The results of the software predictor show the relation of the branches
in the creation and destruction of the heap. The base heapsort results in
Figure \ref{base and memory-tuned heapsorts-1} show that the time taken to
create the heap is not significant relative to the time taken to sort from the
heap.

The most interesting thing is the relation between the predictablility of the
comparisons during the destruction of the 2-, 4- and 8-heaps. Compare the
columns marked `\cc{Destruction, child comparison}' and `\cc{Destruction, child
promotion}' in Figure \ref{base and memory-tuned heapsort-1} (the 2-heap), with
the columns marked `\cc{comparison 1}' to `\cc{child promotion}' in Figure
\ref{base and memory-tuned heapsorts-2} (the 4-heap) and Figures \ref{base and
memory-tuned heapsort-3} and \ref{base and memory-tuned heapsort-4} (the
8-heap).

@@

More importantly, the accuracy of the each successive comparison increases in
the same way as selection sort (see Section \ref{selection branches logn}),
becoming more predictable as the fanout increases. For example, in the 8-heap,
shown in Figures \ref{base and memory-tuned heapsort-3} and \ref{base and
memory-tuned heapsort-4}, the first comparison is predicted correctly 50\% of
the time, while the next is predicted correctly 61\% of the time, then 71\%,
77\%, and so on in a curve, which is shown in Figure \ref{memory-tuned heapsort
curve}.

\begin{figure}
\includegraphics{plots/curve.eps}
\caption{Predictability of branches in memory-tuned heapsort, with a fanout of eight}
\label{memory-tuned heapsort curve}
\end{figure}

### 2005-07-31 docs/tech_report/merge.tex
the grand reaming is almost complete

\section{Double Tiled Mergesort}
\label{double tiled mergesort}
Offsetting the target array to avoid conflicts reduces significantly the number
of level 2 cache misses. However, it does this at the expense of increasing the
number of level 1 cache misses. To avoid this, we further offset the target
array by $Level1CacheSize/2$. It is expected that the number of level 1 misses
should be greatly decreased, at the expense of very few level 2 misses. The
expected improvement is not as great as between base mergesort and tiled
mergesort, but some improvement should be seen.

@@

LaMarca used a 4-heap for his priority queue, and an 8-heap is used here, for
the same reasons as before (see Section \ref{cache heapsort}.). The heap is
aligned, and a cache line is put into the array each time. The smallest key is
at the top of the heap, and this is put into its final position in the array.
The heap is then fixed.

@@

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
pre-sorted, they were pre-sorted to a size four, or eight if the number of
passes was going to be odd. These were reversed for the double multi-mergesort;
since it does a k-way merge into the source array, the sorted cache sized
segments need to be in the auxiliary array. An 8-heap was used for the
multi-mergesorts.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
compensate for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

The cache misses end up being the deciding factor between the performance of
double tiled mergesort and base mergesort, the next best contender. Despite the
reduction in level 2 cache misses, the sharp increase in level 1 misses result
is poor performance for tiled and multi-mergesort. Multi-mergesort is further
encumbered with a much higher instruction count. Double tiled mergesort, with
its low instruction count, and low cache miss rate in both levels 1 and 2,
emerges the winner.

@@

In both double tiled mergesort and base mergesort, approximately two thirds are
of the loops are taken, unlike Knuth's sorts. However, this appears to indicate
an increase in branches, as opposed to a decrease in mispredictions. Overall
this stay is line with the idea that there is a minimum number of branch misses,
below which comparative sorts will not drop.

@@

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversely affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{cache heapsort
curve}). Note that \n{comparison 0} occurs every time, and would be optimized
out by the compiler.

>>>>

\section{Double-aligned Tiled Mergesort}
\label{double-aligned tiled mergesort}
Offsetting the target array to avoid conflicts reduces significantly the number
of level 2 cache misses. However, it does this at the expense of increasing the
number of level 1 cache misses. To avoid this, we further offset the target
array by $Level1CacheSize/2$. It is expected that the number of level 1 misses
should be greatly decreased, at the expense of very few level 2 misses. The
expected improvement is not as great as between base mergesort and tiled
mergesort, but some improvement should be seen.

@@

LaMarca used a 4-heap for his priority queue, and an 8-heap is used here, for
the same reasons as before (see Section \ref{memory-tuned heapsort}.). The heap
is aligned, and a cache line is put into the array each time. The smallest key
is at the top of the heap, and this is put into its final position in the array.
The heap is then fixed.

@@

\section{Results}
\subsection{Test Parameters}
The level 1 cache is 8k, so 1024 keys from each array can be sorted in the level
1 cache. 262144 keys can be sorted in the level 2 cache. When arrays were
pre-sorted, they were pre-sorted to a size four, or eight if the number of
passes was going to be odd. These were reversed for the double-aligned
multi-mergesort; since it does a k-way merge into the source array, the sorted
cache sized segments need to be in the auxiliary array. An 8-heap was used for
the multi-mergesorts.

@@

Despite the improvement in cache misses due to multi-mergesort, it still cannot
compensate for the increase in instruction count as a result, and it has a higher
cycle count on the Pentium 4 (see Figure \ref{merge cycles}) as a result. The
other sorts generally perform in $O(NlogN)$ time, with double-aligned tiled mergesort
being the fastest. Note that the alignment of the mergesort may not work as
desired due to the size of the cache. This was optimized for a 2MB cache, while
the Pentium 4 cache is only 256KB. This means that the two arrays are aligned to
the same cache block. However, since the cache is 8-way associative, this is not
serious, and should not result in a lot of thrashing.

The cache misses end up being the deciding factor between the performance of
double-aligned tiled mergesort and base mergesort, the next best contender.
Despite the reduction in level 2 cache misses, the sharp increase in level 1
misses result is poor performance for tiled and multi-mergesort. Multi-mergesort
is further encumbered with a much higher instruction count. Double tiled
mergesort, with its low instruction count, and low cache miss rate in both
levels 1 and 2, emerges the winner.

@@

In both double-aligned tiled mergesort and base mergesort, approximately two thirds are
of the loops are taken, unlike Knuth's sorts. However, this appears to indicate
an increase in branches, as opposed to a decrease in mispredictions. Overall
this stay is line with the idea that there is a minimum number of branch misses,
below which comparative sorts will not drop.

@@

Overall, when the k-ary heap is used, there are the same number of branch
misses. This means that using the technique to reduce cache misses does not
adversely affect branch prediction rates (it does, however, still affect the
instruction count).  The rate of branch misses is much the same as that in
heapsort, and progresses in the same manner (see Figure \ref{memory-tuned heapsort (8-heap)
curve}). Note that \n{comparison 0} occurs every time, and would be optimized
out by the compiler.

### 2005-07-31 docs/tech_report/ordernsquared.tex
minor changes as rereading
fixing up lots of things
done filename renaming inside files
the grand reaming is almost complete

\section{Selection sort}

@@

\code{Selection sort code}{on2_selectsort.c}
\subsection{Testing}
All elementary sorts are only sorted up to 65536 keys in the simulations, and
262144 keys with the performance counters. Due to the quadratic nature of
these sorts, days and weeks would be required to run tests on the larger arrays.

@@

\section{Insertion sort}

@@

\code{Insertion sort code}{on2_insertsort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick.  Insertion sort
performs in linear time on a sorted list, or slightly unsorted list. With a
random list, there are, on average, there are $N^2/4$ comparisons and $N^2/4$
half-exchanges\footnote{A half-exchange is where only one part of the swap is
completed, with the other value stored in a temporary to be put in place later.}.
As a result, the instruction count should be high, though it is expected to be
lower than selection sort.

@@

\code{Bubblesort code}{on2_bubblesort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick.  Bubblesort has
$N^2/2$ comparisons and exchanges. The number of passes over the array, and size
of each pass is the same as selection sort, and similar instruction count
results are expected.  The extra check to see if the array is sorted adds
$O(NlogN)$ instructions, but this is small in comparison to the number of
instructions saved if the sort ended early.

@@

\section{Bubblesort2}
Bubblesort2 incorporates an improvement to end bubblesort early. The bubblesort
implementation described in the previous section ends early if it detects that
the array is fully sorted.  It keeps a binary flag to indicate if there were any
exchanges along the way.  Instead, bubblesort2 keeps track of the last exchange
made, and starts the next iteration from there. This means that any pockets of
sorted keys are skipped over, not just if they occur at the end. Bubblesort2
code is shown in Figure \vref{Bubblesort2 code}.

\code{Bubblesort2 code}{on2_bubblesort2.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that the bubblesort2 should perform faster than the original in
all metrics. Instruction count, were there no conditions that allowed keys to be
skipped, would be the similar, since few extra instructions are added. The flow
of the algorithm is the same, though there should be fewer iterations,
especially at the start, when iterations are longer, so cache accesses, and
branch misses should be reduced along with instruction count.

@@

\code{Shakersort code}{on2_shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
The instruction count of shakersort is expected to be exactly the same as
bubblesort. The version of shakersort used here, however, does not incorporate
the improvement to end the sort early if the array is already sorted. This will
probably slightly reduce the instruction count as the sort is unlikely to end
early for random keys.

@@

\section{Shakersort2}
Shakersorts incorporates the same improvements over shakersort as bubblesort2
does over bubblesort, that is, it skips over pockets of sorted keys in both
directions, as well as ending early. The code for shakersort2 is in Figure
\vref{Shakersort2 code}.

\code{Shakersort2 code}{on2_shakersort2.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that the improvements will make shakersort faster in all cases
than its original form.

@@

Figure \ref{elementary instructions} shows the instruction count of each
elementary sort. Each sort's instruction count is as predicted. Selection sort
has the most instructions, and insertion sort has approximately half that
number, as predicted. Bubblesort has almost the same number of instructions as
selection sort, but shakersort has a significant number less. The improvement to
bubblesort in bubblesort2 have increased the instruction count, contrary to
predictions, and similarly in the shakersort2. It appears the cost of
skipping over a segment of the array is not worth the additional book-keeping
cost of determining when this can be done.

@@

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse, reducing its number of misses. Shakersort2
is placed roughly between insertion sort and shakersort, showing that ending
early is helpful in its case. This does not appear to be the case with
bubblesort2 though, which performs identically to bubblesort.

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
the branch misses per key.  Branch prediction performance is not surprising.
Each sort has a similar number of branches, with insertion sort having half as
many as selection and bubblesort, and shakersort being half-way between the
two. The number of misses for bubblesort and shakersort are huge; nearly one in
four comparisons are misses. As the data sets gets large, there are nearly
10,000 branch misses involved in sorting a single key. The improved bubblesort
had exactly the same number of misses and branches. The improved shakersort has
a lot less branches, but the same number of misses. This is because skipping
over sorted keys is predictable from the first branch miss.

>>>>

\section{Selection Sort}

@@

\code{Selection sort code}{selectsort.c}
\subsection{Testing}
All elementary sorts are only sorted up to 65536 keys in the simulations, and
262144 keys with the performance counters. Due to the quadratic nature of
these sorts, days and weeks would be required to run tests on the larger arrays.

@@

\section{Insertion Sort}

@@

\code{Insertion sort code}{insertsort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick.  Insertion sort
performs in linear time on a sorted list, or slightly unsorted list. With a
random list, there are, on average, there are $N^2/4$ comparisons and $N^2/4$
half-exchanges\footnote{A half-exchange is where only one part of the swap is
completed, with the other value stored in a temporary to be put in place later.}.
As a result, the instruction count should be high, though it is expected to be
lower than selection sort.

@@

\code{Bubblesort code}{bubblesort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
These instruction count observations are again due to Sedgewick.  Bubblesort has
$N^2/2$ comparisons and exchanges. The number of passes over the array, and size
of each pass is the same as selection sort, and similar instruction count
results are expected.  The extra check to see if the array is sorted adds
$O(NlogN)$ instructions, but this is small in comparison to the number of
instructions saved if the sort ended early.

@@

\section{Improved Bubblesort}
Improved bubblesort incorporates an improvement to end bubblesort early. The
bubblesort implementation described in the previous section ends early if it
detects that the array is fully sorted.  It keeps a binary flag to indicate if
there were any exchanges along the way.  Instead, improved bubblesort keeps
track of the last exchange made, and starts the next iteration from there. This
means that any pockets of sorted keys are skipped over, not just if they occur
at the end. Improved bubblesort code is shown in Figure \vref{Improved bubblesort code}.

\code{Improved bubblesort code}{improved_bubblesort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that improved bubblesort should perform faster than the original
in all metrics. Instruction count, were there no conditions that allowed keys to
be skipped, would be the similar, since few extra instructions are added. The
flow of the algorithm is the same, though there should be fewer iterations,
especially at the start, when iterations are longer, so cache accesses, and
branch misses should be reduced along with instruction count.

@@

\code{Shakersort code}{shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
The instruction count of shakersort is expected to be exactly the same as
bubblesort. The version of shakersort used here, however, does not incorporate
the improvement to end the sort early if the array is already sorted. This will
probably slightly reduce the instruction count as the sort is unlikely to end
early for random keys.

@@

\section{Improved Shakersort}
Shakersorts incorporates the same improvements over shakersort as improved bubblesort
does over bubblesort, that is, it skips over pockets of sorted keys in both
directions, as well as ending early. The code for improved shakersort is in Figure
\vref{Improved shakersort code}.

\code{Improved shakersort code}{improved_shakersort.c}
\subsection{Testing}
\subsubsection{Expected Performance}
It is expected that the improvements will make shakersort faster in all cases
than its original form.

@@

Figure \ref{elementary instructions} shows the instruction count of each
elementary sort. Each sort's instruction count is as predicted. Selection sort
has the most instructions, and insertion sort has approximately half that
number, as predicted. Bubblesort has almost the same number of instructions as
selection sort, but shakersort has a significant number less. The improvement to
bubblesort in improved bubblesort have increased the instruction count, contrary to
predictions, and similarly in the improved shakersort. It appears the cost of
skipping over a segment of the array is not worth the additional book-keeping
cost of determining when this can be done.

@@

As expected, selection sort is the worst. This is due to the fact that it
considers the entire unsorted segment of the array for each key to be put in
place. Bubblesort, with similar behaviour, is only slightly better. Shakersort
has a small amount of temporal reuse, reducing its number of misses. Improved
shakersort is placed roughly between insertion sort and shakersort, showing that
ending early is helpful in its case. This does not appear to be the case with
improved bubblesort though, which performs identically to bubblesort.

@@

Figure \ref{elementary branches} shows the number of branches per key of these
sorts, while Figure \ref{Branch Simulation results for elementary sorts} shows
the branch misses per key.  Branch prediction performance is not surprising.
Each sort has a similar number of branches, with insertion sort having half as
many as selection and bubblesort, and shakersort being half-way between the
two. The number of misses for bubblesort and shakersort are huge.  As the data
sets gets large, there are nearly 10,000 branch misses involved in sorting a
single key. The improved bubblesort had exactly the same number of misses and
branches. The improved shakersort has a lot less branches, but the same number
of misses. This is because skipping over sorted keys is predictable from the
first branch miss.

### 2005-07-31 docs/tech_report/quick.tex
the grand reaming is almost complete

Cache quicksort should have fewer misses than base quicksort as soon as the
array is larger than the cache. It is expected that these misses will make up
for the increase in instruction count.

@@

The instruction count, shown in Figure \ref{quick instructions} is moslty as
predicted. The increase in instructions in both multi-mergesorts is as
predicted, with the binary search version increasing only 25\% against the
sequential search version's 50\% increase. However, the increase we predicted in
the instrcution count of cache quicksort based on the results of both Sedgewick
and LaMarca -  that is, once the efficient insertion sort was replaced with many
small insertion sorts - we observe that the instruction count remains exactly
the same.

@@

The tuned quicksort performed better than the base version, as predicted. Both the
tuned version and the base quicksort both have a significant increase once the
arrays no longer fit in the cache, also as predicted.

The results show again that direct mapped caches can perform better than fully
associative ones. This is true in the case of all the quicksorts, as can be seen
in Figures \ref{base quicksort (no median) level 2 misses} to
\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
in the case of all the quicksorts.

@@

Tuned quicksort had fewer branch misses than base quicksort, due to the fact
that fewer comparitive branches are executed. In base quicksort, we sort the
entire array using insertion sort. In tuned quicksort, we don't sort the pivots,
but only the keys between them.

@@

LaMarca's results are again similar to the results presented here, but not
identical. The shape of the instruction count graph is the same in both cases,
but that of the cache misses is not. The ratio of the cache misses between base
and cache quicksort are the same in both sets of results, but these are lower
than the same misses in LaMarca's results. However, the multi-quicksort results
appear to be the same, though our results are not conclusive either way.

@@

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base, cache and multi quicksorts and base quicksort9}

Firstly, cache quicksort and base quicksort only differ in how they deal with
the insertion. Base quicksort and base quicksort9 have similar numbers of
insertion branches, and a very similar ratio (2/1 correct/incorrect) of
mispredictions. Comparing base quicksort to cahce quicksort (they both use
median of 3), there is actually a decrease in the number of branches handled by
insertion sort, instead of an increase. An increase is expected (thats why
sedgewick moves the insertion to the end), though when you take into account
the sentinel and the outer loop, there probably is an increase in instructions
at least. Base quicksort has 35\% miss rate, but cache has a 34\%. Not a big
difference, but the point is that there is a reduction in branches, but of the
reduced branches, 50\% were unpredictable. I dont think theres anything
interesting in this fact, however.

@@

Note that sequential multi-quicksort has less mispredictions than cache quicksort.

@@

\label{quick-predictors2}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}

@@

We use several variations of quicksort to test this theory. \ref{bently, i
ithkn} suggests partitioning using a median-of-9 instead of a median-of-three. We
created sorts using no median, a median-of-3, and pseudo-medians-of-five, -seven and
-nine. The pseudo-medians used repeated medians-of-three instead of a real
median, as suggestion by \ref{bently, again, i think}. 
%use a cut off

@@

This indicates the same pattern as we saw during heapsort.%
% and mergesort?
The less entropy due to the partitioning key, the less wasted exchanging of keys,
which leads to a general decrease in branches. This decrease means that the
fewer keys are exchanged due to partitioning, leading to less predictable
partitioning. In effect, there are fewer branches pretending to be useful.

The pseudo-median-of-five partitioning performs slightly better than the
median-of-three, and pseudo-median-of-seven performs slightly better again, in
terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
the quicksorts, showing that a pseduo-median-of-seven offers the greatest
tradeoff between the reduced number of branches and instructions, and the
increase in branch midpredictions.

@@

Finally, in multi-quicksort, it may be possible to choose better pivots.
Choosing better (ie, more evenly distributed pivots) would reduce the number of
containers. The advantage of this is that it reduces linearly the number of
pivots to be searched by the sequential search. The most this can be reduced by
is a factor of 3, indicating perfectly distributed pivots. This would reduce the
instruction overhead of the multi-quicksort considerably.

>>>>

Memory-tuned quicksort should have fewer misses than base quicksort as soon as
the array is larger than the cache. It is expected that these misses will make
up for the increase in instruction count.

@@

The instruction count, shown in Figure \ref{quick instructions} is mostly as
predicted. The increase in instructions in both multi-quicksorts is as
predicted, with the binary search version increasing only 25\% against the
sequential search version's 50\% increase. However, the increase we predicted in
the instruction count of memory-tuned quicksort based on the results of both Sedgewick
and LaMarca -  that is, once the efficient insertion sort was replaced with many
small insertion sorts - we observe that the instruction count remains exactly
the same.

@@

The memory-tuned quicksort performed better than the base version, as predicted.
Both the memory-tuned version and the base quicksort both have a significant
increase once the arrays no longer fit in the cache, also as predicted.

The results show again that direct mapped caches can perform better than fully
associative ones. This is true in the case of all the quicksorts, as can be seen
in Figures \ref{base quicksort (median-of-3) (no median) level 2 misses} to
\ref{multi-quicksort (sequential search) level 2 misses} graph that this is true
in the case of all the quicksorts.

@@

Memory-tuned quicksort had fewer branch misses than base quicksort, due to the fact
that fewer comparitive branches are executed. In base quicksort, we sort the
entire array using insertion sort. In tuned quicksort, we don't sort the pivots,
but only the keys between them.

@@

LaMarca's results are again similar to the results presented here, but not
identical. The shape of the instruction count graph is the same in both cases,
but that of the cache misses is not. The ratio of the cache misses between base
and memory-tuned quicksort are the same in both sets of results, but these are lower
than the same misses in LaMarca's results. However, the multi-quicksort results
appear to be the same, though our results are not conclusive either way.

@@

\sixplot{base_quicksort_0.eps}{base_quicksort9_0.eps}{memory_tuned_quicksort_0.eps}{multi_quicksort_0.eps}{sequential_multi_quicksort_0.eps}{}{base,
cache and multi quicksorts and base quicksort (pseudo-median-of-9)}

Firstly, memory-tuned quicksort and base quicksort (median-of-3) only differ in how they deal with
the insertion. Base quicksort (median-of-3) and base quicksort
(pseudo-median-of-9) have similar numbers of insertion branches, and a very
similar ratio (2/1 correct/incorrect) of mispredictions. Comparing base
quicksort to memory-quicksort (they both use
median of 3), there is actually a decrease in the number of branches handled by
insertion sort, instead of an increase. An increase is expected (thats why
sedgewick moves the insertion to the end), though when you take into account
the sentinel and the outer loop, there probably is an increase in instructions
at least. Base quicksort has 35\% miss rate, but cache has a 34\%. Not a big
difference, but the point is that there is a reduction in branches, but of the
reduced branches, 50\% were unpredictable. I dont think theres anything
interesting in this fact, however.

@@

Note that sequential multi-quicksort has less mispredictions than memory-tuned quicksort.

@@

\label{quick-predictors2}
\sixplot{base_quicksort1_0.eps}{base_quicksort_0.eps}{base_quicksort5_0.eps}{base_quicksort7_0.eps}{base_quicksort9_0.eps}{}{base quicksort without a median, with median of 3, and with pseudo-medians of 5, 7 and 9}

@@

We use several variations of quicksort to test this theory. \ref{bently, i
ithkn} suggests partitioning using a median-of-9 instead of a median-of-3. We
created sorts using no median, a median-of-3, and pseudo-medians-of-five, -seven and
-nine. The pseudo-medians used repeated medians-of-three instead of a real
median, as suggestion by \ref{bently, again, i think}. 
%use a cut off

@@

This indicates the same pattern as we saw during heapsort.%
% and mergesort? TODO
The less entropy due to the partitioning key, the less wasted exchanging of keys,
which leads to a general decrease in branches. This decrease means that the
fewer keys are exchanged due to partitioning, leading to less predictable
partitioning. In effect, there are fewer branches pretending to be useful.

The pseudo-median-of-5 partitioning performs slightly better than the
median-of-3, and pseudo-median-of-7 performs slightly better again, in
terms of branch prediction. Figure \ref{TODO} shows the cycle count of each of
the quicksorts, showing that a pseduo-median-of-seven offers the greatest
tradeoff between the reduced number of branches and instructions, and the
increase in branch midpredictions.

@@

Finally, in multi-quicksort, it may be possible to choose better pivots.
Choosing better (ie, more evenly distributed pivots) would reduce the number of
containers. The advantage of this is that it reduces linearly the number of
pivots to be searched by the sequential search. The most this can be reduced by
is a factor of three, indicating perfectly distributed pivots. This would reduce
the instruction overhead of the multi-quicksort considerably.
